Short paper : Automatic predicate testing in formal certification

### 2016-02-20 ProofsAndTests/paper-TAP/content/1-introduction.tex
paper almost finished

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? Too few people realize that the proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show the usual and naive approaches to the problem of a badly specified function which leads to a proof that does not convey the expected guarantee (section~\ref{sect:naiveApproaches})
	\item Show the distinction between structural properties, and the properties on the content, and how to identify what needs to be tested for the former category (section~\ref{sect:differentKindOfProperties}).
	\item Present a way to conduct these tests in the proof assistant itself, by automatically generating terms, and we present a technique to automate these tests as far as possible (section~\ref{sect:testingInside})
\end{itemize}

>>>>

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? Too few people realize that the proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show some basic approaches to the problem of underspecification (section~\ref{sect:naiveApproaches})
	\item Present a way to test in the proof assistant if the predicate effectively captures the notion that we wanted to define, by automatically generating terms, and we automate these tests as far as possible (section~\ref{sect:testingInside})
	\item Present how we can go a step forward by replacing these tests about the predicate by some proofs (section~\ref{sect:aStepForward})
\end{itemize}

### 2016-02-20 ProofsAndTests/paper-TAP/content/3-testingInside.tex
paper almost finished

It is often easier to generate examples of something than it is to precisely define it. We could operate an interesting change of point of view by generating the set of terms that we precisely wanted to describe with the predicate. With our example of sorted lists, that means that we could try to define the same notion of being sorted, but this time with a definition example-based. Writing a function that produces all the sorted lists might be a bit more difficult than just giving a few examples, but this complicated function will have the main advantage of being so different to the predicate that if the two notions agree, then we will know we've gained a great confidence on the predicate.

@@

Since the predicate $isSorted$ is decidable with $decideIsSorted$, we can automatically check whether the generated sorted list of size $n$ are automatically sorted or not by running the decision procedure. But since there can be an infinite amount of generated sorted lists of size $n$ when $n > 0$, we will only check that the generative function and the predicate coincide on a finite unfolding of the resulting lazy list.

in order to ensure that the predicate captures the notion that we want for \emph{any} term, we write a function that will

- it is often easier to generate things rather than to characterise them 
[a quarter of a page : from 3.00 to 3.15

- New : a very nice aspect is that we can now automatically generate \emph{all} the terms that satisfy the predicate. it is therefore possible to see some specific terms and to check weather they should have been captures or not.

- testing the structural properties by automatic generation of the terms that should satisfy them, and running a decision procedure on them.
[half a page : from 3.15 to 3.45]

- Going a step further : ``test automation by proof", which automatises all of them at once, by doing (another) formal proof, which shows that there is equivalence between all the terms generated and the original predicate that we wanted to certify correct according to the original specification.
[half a page : from 3.45 to 4.15]

>>>>

It is often easier to generate examples of something than it is to precisely define it. We can operate an interesting change of point of view by generating the set of terms that we precisely wanted to describe with the predicate. With our example of sorted lists, that means that we need to define the same notion of being sorted, but this time with a definition example-based. Writing a function that produces all the sorted lists might be a bit more difficult than just giving a few examples, but this complicated function will have the main advantage of being so different from the predicate that if the two notions agree, then we will have gained a great confidence on the predicate.

@@

Since the predicate $isSorted$ is decidable with $decideIsSorted$, we can automatically check whether the generated sorted lists of size $n$ are automatically sorted (according to $isSorted$) by running the decision procedure. But since there can be an infinite amount of generated sorted lists of size $n$ when $n > 0$, we will only check that the generative function and the predicate coincide on a finite unfolding of the resulting lazy list. The biggest this unfolding is, the better the guarantee is about $isSorted$.

Let $T$ be a type that only contains three values $A$, $B$ and $C$. Since $T$ is finite, it is therefore obviously recursively enumerable. We define on it the strict order $A < B < C$.
Let's automatically generate the first $m$ sorted lists of $T$, of size $n$, by unfolding $m$ times the result of $generateSortedList$.

\begin{lstlisting}
testGenSorted : (m:Nat) -> (n:Nat) -> Maybe(Vect m (List T))
testGenSorted m n = let x = generateSortedList T TisRecEnu TisOrdered n in
		      	unfold_n_times x m
\end{lstlisting}

We can obtain the first 8 sorted lists of size 5 by evaluating $testGenSorted\ 8\ 5$:

\begin{lstlisting}
Just [[A, A, A, A], [A, A, A, B], [A, A, A, C], [A, A, B, B],
      [A, A, B, C], [A, A, C, C], [A, B, B, B], [A, B, B, C]] : Maybe (Vect 8 (List T))
\end{lstlisting}

Now, instead of simply generating the first $m$ sorted lists, we want to run the decision procedure on all of these $m$ tests in order to know if the predicate and the generating function agree on this portion. The result will be a vector $m$ booleans.

\begin{lstlisting}
testSorted : (m:Nat) -> (n:Nat) -> Vect m Bool
testSorted m n = 
  let x = generateSortedList T TisRecEnu TisOrdered n in
    let y = LLmap (\l => let res = decideIsSorted TisOrdered l in
			    case res of
			      Yes _ => True
			      No _ => False) x in
	unfold_n_times_with_padding y m True
\end{lstlisting}

And we can inspect the result of running the first 8 tests by evaluating $testSorted\ 8\ 5$.

\begin{lstlisting}
[True, True, True, True, True, True, True, True] : Vect 8 Bool
\end{lstlisting}

When we want to test $isSorted$ on a big number of tests, we might not want to inspect by hand the result of each test. We can write a function $testSorted\_result : (m:Nat)\ \rightarrow\ (n:Nat)\ \rightarrow\ Bool$ that calls $testSorted$ and does the $and$ on each element of the resulting vector. Now, we can test the predicate on the first 50 sorted lists of size by running $testSorted\_result\ 50\ 8$ and we get the overall result $True$ which means that the predicate agree with the generator.

### 2016-02-20 ProofsAndTests/paper-TAP/content/4-differentKindOfProperties.tex
paper almost finished

\section{``Structural" VS ``on content" properties}

\label{sect:differentKindOfProperties}

- differences between ``structural" properties, and properties about the ``content".
[half a page : from 2.30 to 3.00]

>>>>

\section{Replacing tests on the predicate by proofs}

\label{sect:aStepForward}

- Going a step further : ``test automation by proof", which automatises all of them at once, by doing (another) formal proof, which shows that there is equivalence between all the terms generated and the original predicate that we wanted to certify correct according to the original specification.
[half a page : from 3.45 to 4.15]
